Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-10-27 demystify.tex
updates

However, no interviews could mean that you will not likely get an RA (\autoref{sec:ra}), which is offered by an individual faculty (if they want you to do research for them, then they likely will interview you first).  In the case of no interviews, your application (and TA/fellowship funding) is decided by the adcom.

@@

\paragraph{Rejection Letters} Schools typically start sending out rejection letters to remaining applicants \emph{after they have finalized their admissions decisions}. Thus, rejection letters are often sent out late (e.g., after April 15th or even much later). Not much you can do here. You can try to contact the school to ask about your status, but they might not reply, they might say they are still reviewing applications, or give you inaccurate information (e.g., you will hear in two weeks). In short, you just have to be patient and wait (however, beware that some schools do not send out rejection letters at all).

@@

\subsectioninfo{Individually faculty member \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so do contact and introduce yourself.}

@@

\subsubsection*{Additional Resources}
\begin{itemize}
  \item \href{https://yonatanbisk.com/emailing_professors.html}{A Note about Emailing Professors} by Yonatan Bisk
  \item \autoref{sec:accept-postpone-decline} How to accept/postpone/decline PhD offers (and do it gracefully)
\end{itemize}

\section{Kiss of Death in Emails}\label{sec:kiss-of-death-emails}

@@

    \item \textbf{Too long}. Keep it to about 3--4 short paragraphs. Less is more and too long emails are often not read and discarded.\\
    Personally I do not want to see course transcript or test scores in the first email. If I am interested I'll ask for them. Attaching CV is OK. Sample papers and links to your Arxiv papers or GitHub projects are also OK if they are relevant to my work.  
    \item \textbf{Flowery greetings and language}. Don't use ``Dear esteemed professor'', don't ``hey there''.  Do not call the prof. by their first name in the first email (some don't care but you don't want to take the risk -- you don't know them that well yet).  Do not use Mr., Mrs., etc. To be safe, use Prof. Lastname or Dr. Lastname (\autoref{sec:address}).

@@

\section{Interpreting Response}\label{sec:interpreting-response}

@@

\chapter{MS Admission}\label{chap:ms}
\chapterinfo{MS focuses on coursework and prepares you for industry, while PhD focuses on research and prepares you for academia or research.}

@@

  \item Objective: an MS is typically to prepare you for industry, while a PhD is to prepare you for research. Some MS has thesis option but the research is much lighter compared to a PhD.

@@

  \item Course requirements: MS has a specific number of courses (typically can be done in 2 years). You graduate with an MS when you're done with the courses. PhD also has coursework requirements (typically taken in your first 2 years) but focuses on research after that.  You graduate with a PhD when you have done enough research and written a dissertation, which usually takes much longer time than coursework.

@@

  \item Funding: MS is typically \emph{not funded}, while PhD is typically funded (\autoref{sec:funding}). See \autoref{sec:ms-funding} for more details on MS funding.

@@

Except for a very few highly selective programs, MS programs are typically much less competitive than PhD programs, i.e., you're likely to get in if you can afford it. Some refer to MS programs as a \emph{cash cow} for the department, as MS students are often not funded and have to pay tuition.

While admission requirements are similar to PhD programs (e.g., GPA, SOP, LORs), research is not a focus in MS programs. In most cases, the main requirements are that you have sufficient background in CS, e.g., through your undergrad degree.

\subsection{Admission Committee}\label{sec:ms-adcom}
MS admission also involves an adcom that reviews applications and makes admission decisions. However, MS admission is typically rolling (i.e., they review applications as they come in and make decisions throughout the admission cycle).  Moreover, unlike PhD that has multiple reviewers for each application, an MS application typically involves only one reviewer and does require much time to review compared to a PhD application (\autoref{sec:ievaluate}).

@@

You typically need to submit the similar materials as in a PhD application, e.g., transcripts, LoRs, SOP, and test scores.  However, research experience is not required, and LoRs can be from anyone who can speak about your academic or working abilities. Moreover, your SOP you should focus on your academic and work background, why you want to get an MS, and how the program fits your goals. As with PhDs, GREs are often not required. Some MS programs do not even require LoRs or SOPs.

@@

\subsection{Funding}\label{sec:ms-funding}
Unlike PhD programs, which often have funding (\autoref{sec:funding}), MS students are typically not funded. The reason is because MS students do not focus on research and thus are not funded through RA and PhD students have priority for TA positions. 
\begin{itemize}
\item (RA) Profs. are not willing to take MS students as RAs because they are not around long enough to be productive. It can take a while for a student to get up to speed and start being productive, and by the time they are, they are about to graduate. Moreover, the goal of most MS students is to get a job, not to do research, so they are not as motivated to do research.
\item (TA) MS students are typically not given TA positions because PhDs are given priority. Many depts. even have shortage of TA positions for PhD students and so they cannot afford to give them to MS students.
\end{itemize}

However, there are always exceptions, e.g., if you have a strong background and can demonstrate that you can be productive, then a prof. might be willing to take you as their RA.
Moreover, some schools have TAs for MS students, and you can apply for these positions. For example, Mason CS dept has quite a few TAs for MS students; my courses in the past two years have MS TAs.

>>>>

However, no interviews mean that you will not likely get an RA (\autoref{sec:ra}), which is offered by an individual faculty (if they want you to do research for them, then they likely will interview you first).  If you have no interviews, your application (and TA/fellowship funding) is decided by the adcom.

@@

\paragraph{Rejection Letters} Schools typically start sending out rejection letters to remaining applicants \emph{after they have finalized their admissions decisions}. Thus, rejection letters are often sent out late (e.g., after April 15th or even much later). Not much you can do here. You can try to contact the school to ask about your status, but they might not reply, they might say they are still reviewing applications, or give you inaccurate information (e.g., you will hear in two weeks). In short, you just have to be patient and wait (and also beware that some schools do not send out rejection letters at all).

@@

\sectioninfo{Individually faculty member \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so do contact and introduce yourself.}

@@

\subsection*{Additional Resources}
\begin{itemize}
  \item \href{https://yonatanbisk.com/emailing_professors.html}{A Note about Emailing Professors} by Yonatan Bisk
  \item \autoref{sec:accept-postpone-decline} How to accept/postpone/decline PhD offers (and do it gracefully)
\end{itemize}

\subsection{Kiss of Death in Emails}\label{sec:kiss-of-death-emails}

@@

    \item \textbf{Too long}. Keep it to about 3--4 short paragraphs. Less is more and too long emails are often not read and discarded. Don't attach course transcript or test scores in the first email. If they are  interested they will ask for them.  Attaching CV is OK. Sample papers and links to your Arxiv papers or GitHub projects are also OK if they are relevant.

    \item \textbf{Flowery greetings and language}. Don't use ``Dear esteemed professor'', don't ``hey there''.  Do not call the prof. by their first name in the first email (some don't care but you don't want to take the risk -- you don't know them that well yet).  Do not use Mr., Mrs., etc. To be safe, use Prof. Lastname or Dr. Lastname (\autoref{sec:address}).

@@

\subsection{Interpreting Response}\label{sec:interpreting-response}

@@

\chapter{MS Admission}\label{chap:ms}
\chapterinfo{MS focuses on coursework and prepares you for \emph{industry}, while PhD focuses on research and prepares you for academia or research.}

@@

  \item Objective: an MS is typically to prepare you for \emph{industry}, while a PhD is to prepare you for research and academia. Some MS has thesis option but the research is much lighter compared to a PhD.

@@

  \item Course requirements: MS has a specific number of courses (typically can be done in 2 years). You graduate with an MS when you're done with the courses. PhD also has coursework requirements (typically taken in your first 2 years as mentioned in \autoref{sec:time}) but focuses on research after that.  You graduate with a PhD when you have done enough research and written a dissertation, which usually takes much longer time than coursework.

@@

  \item Funding: MS is typically \emph{not funded}, while PhD is (\autoref{sec:funding}). See \autoref{sec:ms-funding} for more details on MS funding.

@@

MS programs are much less competitive than PhD programs, i.e., you're likely to get in if you can afford it. Some refer to MS programs as a \emph{cash cow} for the department, as MS students are often not funded and have to pay tuition.

While admission requirements are similar to PhD programs (e.g., GPA, SOP, LORs), research is not a focus in MS programs. In most cases, the main requirements are that you have sufficient background in CS, e.g., through your undergrad degree. This does not mean MS programs are easy to get admitted, but the requirements are much lower compared to PhD programs.

\subsection{Admission Committee}\label{sec:ms-adcom}
MS admission also involves an adcom that reviews applications and makes admission decisions. However, MS admission is typically rolling, i.e., they review applications as they come in and make decisions throughout the admission cycle.  Moreover, unlike PhD that has multiple reviewers for each application, an MS application typically involves only one reviewer and does require much time to review compared to a PhD application (\autoref{sec:ievaluate}).

@@

You typically need to submit the similar materials as in a PhD application, e.g., transcripts, LoRs, SOP, and test scores.  However, research experience is not required, and LoRs can be from anyone who can speak about your academic or working abilities. Moreover, your SOP should focus on your academic and work background, why you want to get an MS, and how the program fits your goals. As with PhDs, GREs are often not required. Some MS programs do not even require LoRs or SOPs.

@@

\subsection{Funding}\label{sec:ms-funding}
Unlike PhD programs, which often have funding (\autoref{sec:funding}), MS students are typically not funded. The reason is because MS students do not focus on research and thus are not funded through RA and PhD students have priority for TA positions. 
\begin{itemize}
\item (RA) Profs. are not willing to take MS students as RAs because they are not around long enough to be productive. It can take a while for a student to get up to speed and start being productive, and by that time, they are about to graduate. Moreover, the goal of most MS students is to get a job, not to do research, so they are not as motivated to do research.

\item (TA) MS students are typically not given TA positions because PhDs are given priority. Many depts. do not have enough TAs for PhD students and so they cannot afford to give them to MS students.
\end{itemize}

However, there are always exceptions, e.g., if you have a strong background and can demonstrate that you can be productive in research, then a prof. might be willing to take you as their RA.
Moreover, some schools have TAs for MS students, and you can apply for these positions. For example, Mason CS has quite a few TAs for MS students; my courses in the past two years have MS TAs.

\section{Selecting and Ranking Schools}\label{sec:selecting-ranking-schools-ms}
%\sectioninfo{How to select and rank MS programs in CS.}

Because of the differences between MS and PhD programs, you should consider different factors when selecting and ranking MS programs. For example, in PhD programs, you should focus on research areas and advisers (\autoref{sec:schoolsandprofs}), while in MS programs, you might want to consider factors such as location, industry connections, and job placement.

\begin{itemize}

\item \textbf{Location:} In addition to living in a place you like (e.g., warm weather), you should also consider the job opportunities in the area.  If you want to work in the tech industry, you might want to consider MS programs in tech hubs like the Bay Area, Seattle, or Boston. For government jobs, you might want to consider schools in the DC DMV area.  Many students at Mason work for the government or defense contractors in the DC area and take classes in the evening or online.

\item \textbf{Industry Connections:} Universities often have strong connections with local companies and can help you get internships and jobs. For example, schools in Washington area often have their students working with Microsoft and Amazon while those in the Bay Area have students working with Google and Facebook.  Mason naturally has strong connections with the government and defense contractors and Amazon, whose 2nd HQ is in Arlington.

\item \textbf{Ranking:} Unlike PhD programs where the adviser and research are perhaps the most important factor, in MS programs the ranking of the school is more important as you do not need to spend 5--7 years doing research.  Thus, you should consider the ranking of the school in CS, especially in the specific area you're interested in.  For example, if you're interested in software engineering, you might want to consider schools with strong SWE programs.

\item \textbf{Living Cost:} As MS students are typically not funded, living cost can be important. You should consider the tuition and living expenses of the area. However, I find that in the beginning living costs, e.g., renting, seems scary but students often find way to make it work, e.g., by sharing an apartment with other students.
